Refunctionalization of Abstracting Abstract Machines (Functional Pearl)

### 2018-03-17 paper.tex
tweak section 3
Editing intro

Abstracting abstract machines (AAM) is a systematic methodology proposed by \citeauthor{van2012systematic}
for constructing abstract interpreters of higher-order functional languages \cite{van2012systematic, van2010abstracting}. 
Given a concrete small-step abstract machine, (e.g, the CESK machine, Krivine's machine or etc), 
by allocating continuations in the store and bound both the value addresses and continuation
addresses to be finite, we obtain an abstract interpreter with finite state space which 
can be used for performing sound static analysis.
One can further instantiate different polyvariant control-flow analysis by using different address allocators \cite{Gilray:2016:ACP:2951913.2951936}.

Applying the same idea on big-step definitional interpreters, \citeauthor{darais2017abstracting} 
build abstract definitional interpreters (ADI) that are written in monadic style \cite{darais2017abstracting}.
One of the advantages of monadic interpreter is modular and composable.
By changing the underlying monads, the definition of interpreter has no change, 
but we can recover different semantics including the concrete semantics and various 
abstract semantics such as context-sensitivity and abstract garbage collection.

Broadly speaking, abstract abstract machines and abstract definitional interpreters are both
abstract interpreters. They are obtained by applying a combination of abstractions to their
concrete counterparts, abstract machines and definitional interperters, perspectively.
In concrete world, the relations among reduction semantics, abstract machines, definitional interpreters,
and monadic interpreters is intensively studied by Danvy and his collaborator in the past years 
\cite{Ager:2003:FCE:888251.888254, Danvy:2001:DW:773184.773202, danvy2004refocusing, Danvy:2008:DIP:1411204.1411206, AGER2004223, ager2005functional, Danvy:2006:RW:2171265.2171268, danvy2009towards, biernacka2009towards}.
The abstract machines implement structural operational semantics in continuation style, 
where the reduction contexts are defunctionalized continuations. One can derive definitional
interpreters by refunctionalizing the evaluation contexts of abstract machines; 
and by defunctionalizing the higher-order functions one may obtain abstract machines in the 
reversed direction.

But in terms of their abstract semantic artifacts, 
the relations between small-step abstract abstract machines and big-step abstract 
definitional interpreters, and the question that how can we interdrive each other is not
well studied.
The fundamental difference between concrete semantics artifacts and abstract semantics 
artifacts is non-determinism. In addition to insure termination, abstract semantics 
artifacts are usually equipped with a cache of explored state that is guaranteed to reach 
the least fixed-point eventually.
In this functional peral, we practically answer the question and relate AAM and ADI 
by presenting a series of syntactical transformations on the program from
small-step abstract abstract machines to big-step abstract definitional interpreters.

In addition, the abstract abstract machines with unbounded stack naturally corresponds 
to abstract definitional interpreters. We show after refunctionalizing an AAM with unbounded
stack and with proper caching algorithm, the pushdown control-flow analysis can be obtained.

@@

\begin{itemize}
\item We show \textbf{linearization} in Section~\ref{linear}. By expressing the 
  non-deterministic choices as first-order data type, we linearize the execution 
  of abstract abstract machines, therefore the transition of machine state becomes deterministic.
    Notably, this introduces anther layer of controls that also will be refunctionalized later.

\item In Section~\ref{fusing}, we then present the \textbf{fusing} transformation. 
  The fusing transformation simply combines the single-step function @step@ and the driving
  function into one, but keeps all the machine state representation.

\item Section~\ref{disen} discusses \textbf{disentangling}, which disassemble the fused AAM to be
  several individual functions, and each function handles one data type represented continuation.

\item \textbf{Refunctionalization} is shown in Section~\ref{refunc} which sequentializes
  the order of abstract execution by higher-order functions.
  For clarity, we first present the refunctionalized AAM without caching. Then we adopt
    a different caching algorithm to guarantee the termination of abstract interpretation.
  In this section, we also review pushdown control-flow analysis and examine how computable 
    and precise call/return match is obtained through these transformations so far.

@@

Now, we obtain a linearized abstract abstract machine. If we image that the classical
AAM explores a graph of reachable states, then the linearized AAM flatten this
graph.

>>>>

Abstracting abstract machines (AAM) is a systematic methodology proposed by~\citeauthor{van2012systematic}
for constructing abstract interpreters of higher-order functional languages~\cite{van2012systematic, van2010abstracting}.
Given a concrete small-step abstract machine, (e.g, the CESK machine, Krivine's machine, etc.),
by allocating continuations in the store and binding both the value addresses and continuation
addresses to be finite, we obtain an abstract interpreter with a finite state space which
can be used for performing sound static analysis.
One can further instantiate different polyvariant control flow analyses by using different address allocators \cite{Gilray:2016:ACP:2951913.2951936}.

Applying the same idea to big-step definitional interpreters, \citeauthor{darais2017abstracting}
built abstract definitional interpreters (ADI) that are written in monadic style \cite{darais2017abstracting}.
One of the advantages of a monadic interpreter is that it is modular and composable.
By changing the underlying monads, the definition of the interpreter is not modified,
but we can recover different semantics, including the concrete semantics and various
abstract semantics such as context-sensitivity and abstract garbage collection.

Broadly speaking, abstract abstract machines and abstract definitional interpreters are both
abstract interpreters. They are obtained by applying a combination of abstractions to their
concrete counterparts, abstract machines and definitional interpreters, respectively.
In the concrete world, the relations among reduction semantics, abstract machines,
definitional interpreters, and monadic interpreters is intensively studied by Danvy and his
collaborators in the past years
\cite{Ager:2003:FCE:888251.888254, Danvy:2001:DW:773184.773202, danvy2004refocusing, Danvy:2008:DIP:1411204.1411206, AGER2004223, ager2005functional, Danvy:2006:RW:2171265.2171268, danvy2009towards, biernacka2009towards}.
These abstract machines implement structural operational semantics in continuation style,
where the reduction contexts are defunctionalized continuations. One can derive definitional
interpreters by refunctionalizing the evaluation contexts of abstract machines,
and by defunctionalizing the higher-order functions, one may obtain abstract machines in the
reverse direction.

But in terms of their abstract semantics artifacts,
the relations between small-step abstract abstract machines and big-step abstract
definitional interpreters, as well as the ability to derive one from the other is not
well-studied.
The fundamental difference between concrete semantics artifacts and abstract semantics
artifacts is nondeterminism. In addition to ensuring termination, abstract semantics
artifacts are usually equipped with a cache of explored state that is guaranteed to reach
the least fixed-point eventually.
In this functional pearl, we practically answer the question and relate AAM and ADI
by presenting a series of syntactical transformations on the program from
small-step abstract abstract machines to big-step abstract definitional interpreters.

In addition, the abstract abstract machines with unbounded stack naturally correspond
to abstract definitional interpreters. We show after refunctionalizing an AAM with unbounded
stack, and with the proper caching algorithm, the pushdown control-flow analysis can be
obtained.

@@

\begin{itemize}
\item We show \textbf{linearization} in Section~\ref{linear}. By expressing the
  nondeterministic choices as a first-order data type, we linearize the execution
  of abstract abstract machines; the transition of machine states therefore becomes deterministic. Notably, this introduces another layer of controls that will also be refunctionalized later.

\item In Section~\ref{fusing}, we then present the \textbf{fusing} transformation.
  The fusing transformation simply combines the single-step function @step@ and the driving
  function into one, but keeps all the machine state representations.

\item Section~\ref{disen} discusses \textbf{disentangling}, which disassembles the fused AAM to be several individual functions, with each function handling one data type represented by a continuation.

\item \textbf{Refunctionalization} is shown in Section~\ref{refunc} which sequentializes
  the order of abstract execution by higher-order functions.
  For clarity, we first present the refunctionalized AAM without caching. We then adopt a different caching algorithm to guarantee the termination of abstract interpretation.
  In this section, we also review pushdown control-flow analysis and examine how computable
    and precise call/return matching is obtained through these transformations.

@@

Now, we obtain a linearized abstract abstract machine. If we image that the classical
AAM explores a graph of reachable states, then the linearized AAM just flattens the
graph to a linear sequence.
